Preprocessing and exploration of the narwhal data for the analysis with smoothSDE

### 2024-05-12 R/preprocessing/data_preprocessing.Rmd
update preprocessing

```{r save points as geojson in data folder}
points_sf <- st_as_sf(dataprep[,c("x","y")], coords = c("x","y"))
st_crs(points_sf)="+init=EPSG:32626 +units=km"
st_write(points_sf,file.path(narwhal_data_path,"narwhals_positions.shp"))
```

@@

## Extract coastline and land polygons

@@

#scoresby sound google
land_google_utm<-st_read(file.path(greenland_data_path,"land_google_utm.shp"))
st_crs(land_google_utm)="+init=EPSG:32626"
```

@@

```{r plot coastline and land}
plot_land=ggplot()+geom_sf(data=land_utm$geometry,fill="grey",col="brown")+
  coord_sf(datum=st_crs(32626))

plot_land_google=ggplot()+geom_sf(data=land_google_utm$geometry,fill="grey",col="blue")+
  coord_sf(datum=st_crs(32626))
```

@@

```{r display coastline , fig.cap="Coastline polygons and linestrings"}
ggplotly(plot_modified_land)
```

```{r}
ggplotly(plot_land_google)
```

@@

```{r summary distances to shore}
summary(res_land$min_distances)
summary(dataprep$DistanceShore)
quantile(res_land$min_distance,probs=c(0.05,0.95))
```

@@

```{r check original points on land}
check_plot1=ggplot()+geom_sf(data=land_google_utm$geometry,fill="grey")+
    coord_sf(datum=st_crs(32626))+
  geom_point(data=dataprep,aes(x,y,col=(DistanceShore<0.05)),size=0.2)+
    scale_color_viridis_d()

@@

df=data.frame(DistanceShore=res_land$min_distances[res_land$min_distances<0.2 & res_land$min_distances>0  ])
histo=ggplot(df, aes(x =DistanceShore)) +
  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Computed distance to shore (km)")

@@

This seems to be pretty uniformly distributed. It is difficult to assess at which distance the narwhals stay from 
the shore in reality.
Let's choose arbitrarily a threshold of 50 metres.

@@

The UTM coordinates for the coastlines are in meters, so that we need to multiply by 1000 when plotting the nearest shore points.

@@

for (id in unique(dataprep$ID)) {
  sub_data=dataprep[dataprep$ID==id,]
  id_track=ggplot()+geom_sf(data=land_utm$geometry,fill=NA)+
    coord_sf(datum=st_crs(32626))+
  geom_point(data=sub_data,aes(px,py),shape=5,size=0.1,col="red")+
  geom_point(data=sub_data,aes(x,y,col=time),size=0.1)+
    geom_path(data=sub_data,mapping=aes(x,y),linewidth=0.1)+
    geom_point(data = sub_data %>% filter(!duplicated(ID)),
             aes(x,y,col=ID), shape = 3, size = 4,col = "red")+
    scale_color_viridis_c()

@@

```{r}
pExpShore=ggplot(dataprep[dataprep$DistanceShore>0.05,],aes(time,ExpShore,col=ID))+ geom_line(color="black")+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  theme(legend.position="none")+xlab("Time (h)")+ylab("Exposure to shore")

>>>>

```{r save points as geojson in data folder}
points_sf <- st_as_sf(dataprep[,c("x","y")], coords = c("x","y"))
st_crs(points_sf)="+init=EPSG:32626 +units=km"
st_write(points_sf,file.path(narwhal_data_path,"narwhals_positions.shp"),append=FALSE)
```

@@

## Extract land polygons to define the border of the domain

@@

@@

```{r plot coastline and land}
plot_land=ggplot()+geom_sf(data=land_utm$geometry,fill="grey",col="brown")+
  coord_sf(datum=st_crs(32626))
```

@@

@@

```{r summary distances to shore}
summary(res_land$min_distances)
summary(dataprep$DistanceShore)
quantile(res_land$min_distance,probs=c(0.05,0.95))
D_low=quantile(res_land$min_distance,probs=0.05)
D_up=median(res_land$min_distance)
```

@@

```{r check original points on land}
check_plot1=ggplot()+geom_sf(data=land_utm$geometry,fill="grey")+
    coord_sf(datum=st_crs(32626))+
  geom_point(data=dataprep,aes(x,y,col=(DistanceShore==0)),size=0.2)+
    scale_color_viridis_d()

@@

df=data.frame(DistanceShore=res_land$min_distances[res_land$min_distances<1 & res_land$min_distances>0  ])
histo=ggplot(df, aes(x =DistanceShore)) +
  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Computed distance to shore (km)")

@@

This seems to be pretty uniformly distributed. It is difficult to assess at which distance the narwhals stay from 
the shore in reality.

@@

@@

for (id in unique(dataprep$ID)) {
  sub_data=dataprep[dataprep$ID==id,]
  id_track=ggplot()+geom_sf(data=land_utm$geometry,fill="grey")+
    coord_sf(datum=st_crs(32626))+
  geom_point(data=sub_data,aes(px,py),shape=5,size=0.1,col="red")+
  geom_point(data=sub_data,aes(x,y,col=time),size=0.1)+
    geom_path(data=sub_data,mapping=aes(x,y),linewidth=0.1)+
    geom_point(data = sub_data %>% filter(!duplicated(ID)),
             aes(x,y,col=ID), shape = 3, size = 4,col = "red")+
    scale_color_viridis_c()

@@

```{r}
df=dataprep
df[df$DistanceShore<D_low,"ExpShore"]=1/D_low
df[df$DistanceShore>D_up,"ExpShore"]=0

pExpShore=ggplot(df,aes(time,ExpShore,col=ID))+ geom_line(color="black")+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  theme(legend.position="none")+xlab("Time (h)")+ylab("Exposure to shore")
